---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### Loading and preprocessing the data
```{r}
if(!file.exists("activity.csv")) {
    unzip("activity.zip")
}
act <- read.csv("activity.csv")
```

Convert the date variable into the Date format:

```{r}
act$date <- as.Date(act$date, "%Y-%m-%d")
```

### What is the mean total number of steps taken per day?

Calculate total number of steps taken per day and plot it in a histogram:

```{r}
steps_per_day <- tapply(act$steps, act$date, sum, na.rm=TRUE)
hist(steps_per_day)
```

```{r}
paste("Mean of steps per day:", mean(steps_per_day, na.rm = TRUE))
paste("Median of steps per day:", median(steps_per_day, na.rm = TRUE))
```

### What is the average daily activity pattern?

```{r, echo=TRUE, message=FALSE}
library(dplyr)
```

Calculate the average steps for each 5-min time interval across all days:

```{r}
steps_by_interval <- act %>%
                     group_by(interval) %>% 
                     summarize(avg = mean(steps, na.rm = TRUE))
```

Make a time-series plot of these average steps:

```{r}
plot(steps_by_interval, type="l", xlab="Interval", ylab="Average steps",
     main="Average steps for each 5-min time interval")
```

Find 5-minute interval containing the maximum number of steps:

```{r}
steps_by_interval[which.max(steps_by_interval$avg), ]
```

### Imputing missing values
 
Calculate total numbers of missing values in the dataset:

```{r}
sum(is.na(act))
```

Create new dataset by filling the missing values with the average of the corresponding 5-minute interval:

```{r}
act_fill <- act %>% transform(
    steps = ifelse(
        is.na(steps),
        yes = subset(steps_by_interval, interval==interval)$avg,
        no = steps
        )
    )
```

Test if the new dataset really has no more missing values:

```{r}
sum(is.na(act_fill))
```

Make a histogram of the total number of steps taken each day for the new data frame:

```{r}
steps_per_day_fill <- tapply(act_fill$steps, act_fill$date, sum)
hist(steps_per_day_fill)
```

```{r}
paste("Mean of steps per day:", mean(steps_per_day_fill))
paste("Median of steps per day:", median(steps_per_day_fill))
```
The values are slightly higher than before the NA values were filled.

### Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day:

```{r}
act_fill <- act_fill %>% mutate(
    week_cat = factor(ifelse(weekdays(date) %in% c("Samstag", "Sonntag"),
                      yes = "weekend", no = "weekday")
               )
    )
```

Make a time-series plot comparing average daily steps between weekdays and weekends:

```{r, echo=TRUE, message=FALSE}
library(ggplot2)
```

```{r, message=FALSE}
# calculate aggregated steps
steps_by_interval_fill <- act_fill %>%
                     group_by(week_cat, interval) %>% 
                     summarize(avg = mean(steps))
# create plot
ggplot(steps_by_interval_fill, aes(x=interval, y=avg, color=week_cat)) +
    geom_line() + 
    labs(title="Average daily steps by interval", 
         x="Interval", y="Average Steps") +
    facet_wrap(~week_cat, nrow = 2)

```

